% arara: xelatex
\documentclass[12pt]{article}

\usepackage{physics}


\usepackage{tikz} % картинки в tikz
\usepackage{microtype} % свешивание пунктуации

\usepackage{array} % для столбцов фиксированной ширины

\usepackage{indentfirst} % отступ в первом параграфе

\usepackage{sectsty} % для центрирования названий частей
\allsectionsfont{\centering}

\usepackage{amsmath, amsfonts, amssymb} % куча стандартных математических плюшек

\usepackage{comment}

\usepackage[top=2cm, left=1.2cm, right=1.2cm, bottom=2cm]{geometry} % размер текста на странице

\usepackage{lastpage} % чтобы узнать номер последней страницы

\usepackage{enumitem} % дополнительные плюшки для списков
%  например \begin{enumerate}[resume] позволяет продолжить нумерацию в новом списке
\usepackage{caption}

\usepackage{url} % to use \url{link to web}

\usepackage{fancyhdr} % весёлые колонтитулы
\pagestyle{fancy}
\lhead{Time Series and Stochastic Processes}
\chead{}
\rhead{Midterm retake, 2021-12-06}
\lfoot{2021-2022}
\cfoot{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

\usepackage{tcolorbox} % рамочки!

\usepackage{todonotes} % для вставки в документ заметок о том, что осталось сделать
% \todo{Здесь надо коэффициенты исправить}
% \missingfigure{Здесь будет Последний день Помпеи}
% \listoftodos - печатает все поставленные \todo'шки


% более красивые таблицы
\usepackage{booktabs}
% заповеди из докупентации:
% 1. Не используйте вертикальные линни
% 2. Не используйте двойные линии
% 3. Единицы измерения - в шапку таблицы
% 4. Не сокращайте .1 вместо 0.1
% 5. Повторяющееся значение повторяйте, а не говорите "то же"



\usepackage{fontspec}
\usepackage{polyglossia}

\setmainlanguage{english}
\setotherlanguages{english}

% download "Linux Libertine" fonts:
% http://www.linuxlibertine.org/index.php?id=91&L=1
\setmainfont{Linux Libertine O} % or Helvetica, Arial, Cambria
% why do we need \newfontfamily:
% http://tex.stackexchange.com/questions/91507/
\newfontfamily{\cyrillicfonttt}{Linux Libertine O}

%\AddEnumerateCounter{\asbuk}{\russian@alph}{щ} % для списков с русскими буквами
%\setlist[enumerate, 2]{label=\asbuk*),ref=\asbuk*}

%% эконометрические сокращения
\DeclareMathOperator{\Cov}{\mathbb{C}ov}
\DeclareMathOperator{\Corr}{\mathbb{C}orr}
\DeclareMathOperator{\Var}{\mathbb{V}ar}

\let\P\relax
\DeclareMathOperator{\P}{\mathbb{P}}

\DeclareMathOperator{\E}{\mathbb{E}}
% \DeclareMathOperator{\tr}{trace}
\DeclareMathOperator{\card}{card}
\DeclareMathOperator{\plim}{plim}
\DeclareMathOperator{\pCorr}{\mathrm{p}\mathbb{C}\mathrm{orr}}


\newcommand \hb{\hat{\beta}}
\newcommand \hs{\hat{\sigma}}
\newcommand \htheta{\hat{\theta}}
\newcommand \s{\sigma}
\newcommand \hy{\hat{y}}
\newcommand \hY{\hat{Y}}
\newcommand \e{\varepsilon}
\newcommand \he{\hat{\e}}
\newcommand \z{z}
\newcommand \hVar{\widehat{\Var}}
\newcommand \hCorr{\widehat{\Corr}}
\newcommand \hCov{\widehat{\Cov}}
\newcommand \cN{\mathcal{N}}
\newcommand \RR{\mathbb{R}}
\newcommand \NN{\mathbb{N}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\cH}{\mathcal{H}}


\begin{document}

Short rules: 120 minutes, online without proctoring. You may use any source you want but don't cheat.
\vspace{10pt}

\textbf{Pledge:}
\vspace{10pt}
\begin{tcolorbox}
Start exam by writing the following honor pledge and signing it.

\vspace{10pt}

I pledge on my honor that I will not give nor receive any 
unauthorized assistance on this exam.
\end{tcolorbox}
\vspace{20pt}

\textbf{Problems:}


\begin{enumerate}

\item (10 points) Consider the Markov chain with the transition matrix
\[
  P = \begin{pmatrix}
    0.2 & 0.2 & 0 & 0.6 & 0 \\
    0.3 & 0.3 & 0.4 & 0 & 0\\
    0 & 0 & 0.3 & 0.7 & 0 \\
    0 & 0 & 0.8 & 0.2 & 0 \\
    0 & 0 & 0 & 0 & 1 \\
  \end{pmatrix}.
\]

\begin{enumerate}
  \item (3 points) Split the chain in classes and classify them into closed or not closed.
  \item (2 points) Classify the states into recurrent or transient.
  \item (5 points) A Hedgehog starts in the state one and moves 
  randomly between states according to the transition matrix.

  What is the approximate probability that the Hedgehog will be in the 
  state four after $10^{2021}$ moves?
\end{enumerate}

Note: state number is the row (or column) number.

  \item (10 points) Consider infinite ladder with steps numbered from $0$ to infinity. 
  I start at step $0$. Every day with probability $u$ I go one step up.
  With probability $d$ I go one step down. With probability $1-u-d$ I stay on the same step.

  If I am at step $0$ then I stay there with probability $1-u$ because it's impossible to go down. 

  Consider the case $d>u$. 
  
  What is the probability that I will be at step $0$ after $10^{1000}$ days?

  \item (10 points) The random variables $X_i$ are independend and uniformly distributed on $[0;2]$.
  Find the probability limit
\[
\plim_{n\to\infty}  \max \left\{ \frac{\sum_{i=1}^{10} X_i}{n}, \frac{\sum_{i=1}^n X^3_i}{n+1} \right\}.
\]


\item (10 points) Taxis arrive to the station according to the Poisson process with rate 1 per 5 minutes. 

Let $Y_t$ be the number of taxis that will arrive between 0 and $t$ minutes.

\begin{enumerate}
  \item (5 points) Sketch the probability $\P(Y_{t+3} = 1 \mid Y_t = 0)$ as a function of $t$.
  \item (5 points) Sketch the covariance $\Cov(Y_t, Y_{60})$ as a function of $t$.
\end{enumerate}

Note: special points like intercepts or extrema should be explicitely marked.

\newpage 
\item (10 points) The moment generating function of a random variable $X$ is $1/(1-2t)$.
\begin{enumerate}
    \item Find the moment generating function of $2X$.
    \item Find the moment generating function of $X + Y$ where $X$ and $Y$ are independent and identically distributed.
    \item Do you remember the sum of geometric progression? Find $\E(X^{2021})$.
\end{enumerate}

\item (20 points) Variables $X_1$, $X_2$, \ldots $X_{100}$ are independent and identically distributed
with mean $1$ and variance $2$. Each $X_i$ has only three possible values: 0, 1, and 2. 

\begin{enumerate}
  \item (5 points) How many events are in sigma-algebras $\sigma(X_1, X_2)$ and $\sigma(X_1 - X_2)$?
  \item (5 points) If possible provide an example of events $A$ and $B$ such that: $A\in \sigma(X_1, X_2)$ but $A\not\in\sigma(X_1 - X_2)$; $B\in \sigma(X_1 - X_2)$ but $B\not\in\sigma(X_1, X_2)$.
  \item (10 points) Find $\E(X_1 + \ldots + X_{100} \mid X_1 + \ldots + X_{50})$ and $\E(X_1 + \ldots + X_{50} \mid X_1 + \ldots + X_{100})$.
\end{enumerate}






\end{enumerate}


\end{document}

